### 2022-11-11 submission/manuscript.tex
First full draft of Introduction

\textbf{\emph{\url{https://edepot.wur.nl/547087}.}} ``Log- ratio PCA is
designed to give results that are library size- independent. However, as
we demonstrated mathematically and with examples based on simulated and
real data, log- ratio PCA becomes library size- dependent, if there are
many infrequent taxa (many zeroes) and library sizes differ largely. In
this situation, the row centring used in log- ratio PCA brings an effect
of r (the row mean of the log- transformed counts) in the clr-
transformed matrix. Note that this effect is irrespective of whether or
not these infrequent taxa are genuine or due to sequencing noise or
allocation error. This library size dependence is unexpected in the
sense that, after applying the clr, the transformed matrix is free of
the effect of the row totals for strictly positive data
(yij\textgreater0 for all i and j). We additionally demonstrate that
library size variability causes a loss in power in dextecting an effect
of x with log- ratio RDA. If there is additionally a correlation between
treatment and the library size, the type 1 error for detecting the
effect of x can be seriously inflated.''

\textbf{\emph{\url{https://www.nature.com/articles/s41522-020-00160-w}.}}
``An important characteristic of a feature table is that it is typically
sparse, sometimes as many as \textasciitilde90\% are zero entries21,
which creates a challenge for analyzing rare taxa. A quick and simple
strategy to deal with excess zeros is to add a small positive constant
(e.g.~1) called pseudo-count14,22 to each cell of the feature table. The
addition of a pseudo-count becomes necessary when using methods of
analysis that require log transformation of the observed counts. Even
though adding a pseudo-count is simple and widely used, the choice of
the pseudo-count is ad hoc. Studies have shown that differential
abundance or clustering results could be sensitive to the choice of
pseudo count23,24. Although different values of pseudo counts have been
discussed in the literature23,24,25,26, to the best of our knowledge,
there is no consensus on how to choose the optimal value. Other
strategies involve modeling zero counts by some probability
models21,27.''

\url{https://academic.oup.com/bioinformatics/article/34/16/2870/4956011}
\url{https://www.ncbi.nlm.nih.gov/pmc/articles/PMC6755255/}

\begin{itemize}
\tightlist
\item
  Motivation

  \begin{itemize}
  \tightlist
  \item
    Problem of uneven sampling effort
  \end{itemize}
\item
  What is rarefaction? History, reason for rarefaction

  \begin{itemize}
  \tightlist
  \item
    Repeated down sampling of datasets to a common number of
    observations to calculate the average value to ascertain the
    expected value of a metric for the metric under study; typically
    richness
  \item
    Control for unneven sampling effort
  \item
    Methods vary in their sensitivity to uneven sampling
  \item
    Compositional data analysis
  \end{itemize}
\item
  Reasons behind ``rarefaction is inadmissable''

  \begin{itemize}
  \tightlist
  \item
    Weird simulation
  \end{itemize}
\item
  Alternative approaches and claims

  \begin{itemize}
  \tightlist
  \item
    sampling invariance
  \end{itemize}
\item
  Goal of this study
\end{itemize}

This analysis included 16S rRNA gene sequence data from from 12 studies
that characterized the variation in bacterial communities from diverse
environments (Table 1). The original studies generated the sequence data
by pooling separate PCR products that were generated by amplifying the
V4 region of the 16S rRNA gene from the bacterial DNA in multiple
samples. Because pooling equimolar quantities of DNA is frought with
difficulties, it was common to observe wide variation in the number of
sequences in each sample (Table 1 and Figure S1).

@@

Over the past decade, the question of whether to rarefy microbial
community sequence data has become controversial. The analyses I
presented here strongly indicate that rarefaction is necessary to
control for uneven sampling effort when comparing communities based on
alpha and beta diversity indices. Compared to all other methods,
rarefaction removed the correlation between sequencing depth and alpha
or beta diversity metrics when the sequencing depth varied by as much as
97-fold across samples. I showed that this correlation could lead
falsely detecting differences between treatment groups if sampling depth
and sequencing effort are confounded (e.g., 1). The correlation with
sampling effort leads to an artificial increase in the variation between
samples and a reduced power to detect true differences in alpha and beta
diversity. For these reasons, rarefaction is a valuable tool to control
for uneven sampling effort until improved statistical procedures are
developed or it becomes possible to more evenly distribute sequencing
effort across samples.

The primary alternative to rarefaction that many advise is to estimate
alpha diversity has been to use non-parametric or parametric methods
using raw data and to then compare the estimates (2). My data
demonstrates that such approaches are limited for several reasons.
First, it has been widely observed that non-parametric richness esimates
such as ACE and Chao1 are sensitive to sampling effort. Therefore, these
strategies do not, in practice, remove the effects of sampling effort.
Second, parametric approaches, such as those implemented in the
breakaway R package, generate confidence intervals that are likely to
include the true richness and theoretically shrink with increased
sampling effort. For most samples, the confidence intervals around the
estimates are too wide to be meaningful, again leading to an inability
to remvoe the effects of sampling effort. Third, it has become an
increasingly common practice for researchers to remove sequences that
are rare in a sample (e.g., those that appear once). Although that
approach was not taken in this study, removing rare sequences would skew
the distribtion of sequences and OTUs leading to a distortion of the
measurement of alpha diversity (3). The effects of removing rare
sequences would vary across samples depending on the number of sequences
in each sample. One interesting result of this analysis was the
demonstration that as metrics that depend less on rare taxa are used,
the effect of uneven sampling effort was reduced. For example, richness
counts a sequence appearing once as much as sequence appearing 1,000
times, while Shannon diversityindex places more emphasis on the more
abundant sequence, and the inverse Simpson index even more. Although
normalizing communities to a common number of sequences is also
suggested (e.g., SRS normalization) to control for uneven sampling
effort, the current analysis indicates that its performance does not
meet that of rarefaction. For analysis of alpha diversity metrics,
rarefaction is the most effective and consistent approach to control fo
uneven sampling effort.

Use of relative abundances, normalized counts, variance stabilizing
transformations, and centered log ratios have each been recommended as
more robust alternatives to rarefaction. Again, the only approach that
consistently removed the effects of uneven sampling effort was
rarefaction. Most of the recommendations borrow techniques from methods
for identifying differential gene expression. Unfortunately, there is an
important but underappreciated difference between gene expression and
community data. This is the interpretation of unobserved data. For gene
expression analysis in a single organism the lack of any sequencing data
for a gene would indicate that its expression was below the limit of
detection. Sequencing the same organism under multiple conditions would
not lead to a seemingly unbouned number of genes in the organims.
Rather, the number of genes has a definite limit that is knowable from
the genome sequence. With microbiome data, an unobserved sequence could
mean that the organism was present, but below the limit of detection or
that the organism was missing. Because we have yet to exhaustively
sample any community in the same way we have sequenced a single genome
it is unreliable to impute the presence of all organisms. Yet, this is
exactly what the variance stabilization transformation and most CLR
techniques do. This analysis has demonstrated a clear correlation
between distances calculated by these methods and sequencing effort.
This result is at odds with the claims by others that the distances are
scale invariant (4). Again, rarefaction is the most effective and
consistent approach to control for uneven sampling effort when
calculating beta diversity metrics.

Two common critique of rarefaction is that the approach ommits valid
data and that the selection of the value to rarefy the sequence data to
is arbitrary (5). I disagree with both critiques. All of the data are
used to calculate the mean value of the rarefied metrics; none of it is
excluded. When the dataset is subsampled, every sequence has a random
chance of being included in the calculated metric. When that subsampling
is repeated a large number of times (e.g., 100 or 1000) the risks of
ignoring or oversampling rare taxa are mititgated. Furthermore, it is
curious that the study making the original critique removed sequences
that did not appear in at least three samples or had an abundance of one
in any sample. A parallel analysis to this study has demonstrated that
many of these sequences are likely valid and that removal of rare
sequences can bias alpha and beta diversity metrics and reduce
statistical power (3). As for the second critism, I would resist the
claim that the selection of the rarefaction threshold is arbitrary. In
practice, there is a tradeoff between sampling breadth and depth.
Greater breadth will increase the statistical power to compare treatment
groups and greater depth will increase the resolution to describe the
communities. My process for picking a break involves looking for a
natural break in the distribution of the number of sequences. For
example, the Lake dataset used in this study had a clear break around
10,000 sequences. I would also consider what samples are below any break
that I select. If there were critical samples below the break I would
either reduce the break point or I would generate more sequences from
those samples. Furthermore, regardless of the break point one selects,
they could repeat an analysis at multiple thresholds to assess the
tradeoff between statistical power and resolution. As shown in my
analysis of Good's coverage values, most studies obtain an ample level
of coverage and would need to increase their sampling depth by 10 fold
to increase the coverage by several percentage points. I contend that
for most studies increased sequencing effort should be applied to
improve sampling breadth and statistical power over depth and
resolution.

The up to 100-fold difference in sample sizes is an unfortunate
byproduct of how seqencing libraries are constructed. Researchers
perform separate PCRs for each sample with unique index (aka, barcode)
sequencs that all them to later identify which sample each sequence
belongs. When the PCRs are pooled efforts are taken to pool the
fragments in equimolar ratios. Researchers use one of two approaches.
First, they often will quantify the concentration of DNA from each PCR
and then pool DNA in the desired amounts. Alternatively, they may use
normalization plates where each well can hybridize a uniform amount of
DNA that is then eluted and pooled. Clearly, both approaches have
limitations that reduce the ability to truly achieve equimolar mixture.
In addition, for some samples it is common to co-amplify non-specific
DNAs which may add to the challenges of achieving equimolar mixtures of
the desired amplicons (1). Regardless, it is clear that better
strategies are needed to reduce the variation in the number of sequences
generated for each sample.

@@

\textbf{Choice of datasets.} The specific studies used in this study
were selected because their data was publicly accessible through the
Sequence Read Archive, the original investigators sequenced the V4
region of the 16S rRNA gene using paired 250 nt reads, and my previous
familiarity with the data. The use of paired 250 nt reads to sequence
the V4 region resulted in a near complete two-fold overap of the V4
region resulting in high quality contigs with a low sequencing error
rate (6). These data were processed through a standardized sequence
curation pipeline to generate operational taxonomic units (OTUs) using
the mothur software package (6, 7). OTUs were identified using the
OptiClust algorithm to cluster sequences together that were not more
than 3\% different from each other (8).

\textbf{Null community model.} Null community models were generated such
that within a dataset the number of sequences per sample and the number
of sequences per OTU across all samples within the dataset were the same
as was observed in original. This model effectively generated
statistical samples of a single community so that there should not have
been a statistical difference between the samples. This model
implemented by randomly assigning each sequence in the dataset to an OTU
and sample while keeping constant the number of sequences per sample and
the total number of sequences in each OTU. This is a similar approach to
that of the IS algorithm described by Ulrich and Gotelli (9). Because
the construction of the null models was a stochastic process, 100
replicates were geneated for each dataset.

@@

\textbf{Test of statistical significance.} Statistical comparisons of
alpha diversity metrics across the simulated treatment groups were
performed using the non-parametric two-sample Wilcoxon test as
implemented in R. This test was selected because the alpha-diversity
metrics tended to not be normally distributed and each dataset required
a different transformation to normalize the data. Comparisons of beta
diversity metrics were performed using the adonis2 function from the
vegan (v.2.6.2) R package (10). The adonis2 function implements a
non-parametric multivariate analysis of variance using distances
matrices as described by McArdle and Anderson (11). Throughout this
study I used 0.05 as the threshold for asssessing statistical
significance.

@@

\textbf{Alpha diversity calculations.} Various strategies for handling
uneven sampling effort were evaluated to identify the best approach for
calculating community richness and Shannon and inverse Simpson diversity
indeices. Raw OTU counts were used as input to calculate sample richness
and Shannon and inverse Simpson diversity using mothur (7, 12). Shannon
diversity was calculated as

@@

The inverse Simpson diversity was calculated as \(1/D_{simpson}\). In
both formulae, \(n_i\) is the number of sequences in OTU \(i\) and \(N\)
is the number of sequences in the sample. Rarefaction of richness,
Shannon diversity and Inverse Simpson diversity values were carried out
in mothur. Briefly, mothur calculates each value on a random draw of the
same number of sequences from each sample and obtains a mean value based
on 1,000 random draws. Scaled ranked subsampling (SRS) was used to
normalize OTU counts to the size of the smallest sample in each dataset
using the SRS R package (v.0.2.3)(13). Normalized OTU counts were used
to calculate sample richness and Shannon and inverse Simpson diversity
values using mothur. Data normalized by cumulative sum scaling (CSS)
were not reported for alpha-diversity values since the relative
abundances of the features do not change with the normalization
procedure (14). The non-parametric bias-corrected Chao1 and ACE richness
estimators (15) and a non-parametric estimator of the Shannon diversity
(16) were calculated using raw OTU counts with mothur. Parametric
estimates of sample richness were calculated using the breakaway (BA) R
package (v.4.7.9)(17). The current analysis reports both the results
from running default model selection procedure and the Poisson model.
The default model selection returned either the Kemp, Negative Binomial,
or Poisson models. Relative abundance data were not used to calculate
alpha diversity metrics since the richness and evenness does not change
from the raw data when dividing each sample by the total number of
sequences in the sample.

\textbf{Beta diversity calculations.} Similar to the alpha diversity
calculations, multiple approaches were used to control for uneven
sampling effort and calculate beta diversity. Raw and OTU counts were
used for input to calculate the Jaccard, Bray-Curtis, and Euclidean
dissimilarity indices using the vegdist function from the vegan R
package (v.2.6.2)(10). The Jaccard index was calculated as

@@

Three approaches were taken to normalize the number of sequences across
samples within a dataset. Scaled ranked subsampling (SRS) and cumulative
sum scaling (CSS) were used to normalize raw OTU counts using the SRS
(v.0.2.3) and metagenomeSeq (v.1.36.0) R packages, respectively (13,
14). The normalized counts were then used to calculate Jaccard and
Bray-Curtis dissimilarity indices. Finally, the variance-stabilization
transformation (VST) as implemented in the DESeq2 (v.1.34.0) R package
was used to normalize the data as described by McMurdie and Holmes (5,
18). Because the VST approach generates negative values, which are
incompatible with calculating Jaccard and Bray-Curtis dissimilarity
values, Euclidean distances were calcualted instead.

@@

where \$x\_\{ij\} was the number of sequences observed for OTU \(i\) in
sample \(j\) and \(g()\) was the geometric mean \(x_j\) was the counts
of the \(S_T\) OTUs in sample \(j\). Because the geometric mean is zero
if any OTU is absent from a sample, the CLR is undefined when there are
unobserved OTUs in a sample. To overcome this problem, I attempted a
four approaches. The first, Zero CLR, imputed the value of the zeroes
based on the observed data using the zCompositions (v.1.4.0.1) R package
(19). The second, One CLR, added a pseudocount of 1 to the abundance of
all OTUs (14, 20). The third, Nudge CLR, added a pseudocount of 1
divided by the total number of sequences in a sample to each OTU in the
sample (20, 21). The final approach, Robust CLR, removed unobserved OTUs
prior to calcualting the CLR (4).

@@

\hypertarget{refs}{}
\begin{CSLReferences}{0}{1}
\leavevmode\vadjust pre{\hypertarget{ref-Morris2013}{}}%
\CSLLeftMargin{1. }%
\CSLRightInline{\textbf{Morris A}, \textbf{Beck JM}, \textbf{Schloss
PD}, \textbf{Campbell TB}, \textbf{Crothers K}, \textbf{Curtis JL},
\textbf{Flores SC}, \textbf{Fontenot AP}, \textbf{Ghedin E},
\textbf{Huang L}, \textbf{Jablonski K}, \textbf{Kleerup E},
\textbf{Lynch SV}, \textbf{Sodergren E}, \textbf{Twigg H}, \textbf{Young
VB}, \textbf{Bassis CM}, \textbf{Venkataraman A}, \textbf{Schmidt TM},
\textbf{Weinstock GM}. 2013. Comparison of the respiratory microbiome in
healthy nonsmokers and smokers. American Journal of Respiratory and
Critical Care Medicine \textbf{187}:1067--1075.
doi:\href{https://doi.org/10.1164/rccm.201210-1913oc}{10.1164/rccm.201210-1913oc}.}

\leavevmode\vadjust pre{\hypertarget{ref-Willis2019}{}}%
\CSLLeftMargin{2. }%
\CSLRightInline{\textbf{Willis AD}. 2019. Rarefaction, alpha diversity,
and statistics. Frontiers in Microbiology \textbf{10}.
doi:\href{https://doi.org/10.3389/fmicb.2019.02407}{10.3389/fmicb.2019.02407}.}

\leavevmode\vadjust pre{\hypertarget{ref-Schloss2020}{}}%
\CSLLeftMargin{3. }%
\CSLRightInline{\textbf{Schloss PD}. 2020. Removal of rare amplicon
sequence variants from 16S {rRNA} gene sequence surveys biases the
interpretation of community structure data.
doi:\href{https://doi.org/10.1101/2020.12.11.422279}{10.1101/2020.12.11.422279}.}

>>>>

The ability to generate millions sequence reads from 16S rRNA gene
amplicons and metagenomic datasets has allowed researchers to multiplex
multiple samples on the same sequencing run by pooling separate PCRs
that can be deconvoluted later based on index (aka barcode) sequences
that embeded into the primer sequence (1, 2). Unfortunately, it is
common to observe variation in the number of sequence reads from each
sample vary by as much as 100-fold (e.g, see Figure S1). This occurs
because pooling of DNA from multiple PCRs is frought with numerous
opportunities for technichal errors to compound leading to a skewed
distribution. Aside from developing better methods of pooling DNA, the
question of how to control for uneven sampling effort in microbial
ecology studies has become controversial.

The practice of rarefaction has been commonly used in ecology for more
than 50 years as a tool to control for uneven sampling effort across
experimental replicates (3, 4). Microbial ecologists have used it to
compare 16S rRNA gene sequence data for the past 25 years (5--7). When
data are rarefied, the investigator selects a desired threshold of
sampling effort and removes any samples below that threshold. They then
randomly select that many sequences, with replacement from each sample.
Based on the obseved sequence counts, the researcher can then calcuate
alpha diversity metrics including richness and diversity indices or beta
divesity metrics such as a Jaccard or Bray-Curtis dissimilarity index. I
refer to this single sampling as a subsample; this method is implemented
as the sub.sample function in mothur (8) and the rrarefy function in the
vegan R package (9). Rarefaction repeats the subsampling a large number
of times (e.g., 100 or 1,000) and calculates the mean of the alpha or
beta diersity metric over those subsamplings; rarefaction is implemented
in mothur using the summary.single and dist.shared functions (8) and
with the vegan R package using the rarefy or avgdist fuctions (9).
Rarefaction effectively tells a researcher what an alpha or beta
diversity metric would have been for a collection of samples if they
were all evenly sampled. Although a closed form equation exists to
calcualte the expected richness (4), it is computationally easier to
empirically calculate rarefied richness and other alpha and beta
diversity metrics.

In 2014, McMurdie and Holmes (10) announced that rarefaction of
microbial community data is ``statistically inadmissible'' because it
omits valid data. In their simulations, they observed that the use of
rarefaction reduced the statistical power to compare communities based
on beta diversity metrics. The communities used in this study were
simulated by mixing operational taxonomic unit (OTU) frequencies from
marine and human fecal communities, which were then sampled to generate
40 replicate communities each sampled to different depths. The
simulation with the greatest depth of coverage obtained an average of
10,000 sequences per sample. Analysis of the R code that they used
indicates that the simulations were actually only based on a single
subsample and not truly rarefied data. Furthermore, the authors removed
any OTU from the simulated communities that was not observed in three or
more samples. This is a frequently used approach for analyzing
microbiome data (11). Unfortunately, it removes valid data, skews the
distribution of the community, and removes true patchiness from the
representation of the communities (12). Finally, the analysis did not
include analysis of alpha diersity metrics and chose to assess beta
diversity metrics based on clustering quality of samples rather than
using the more frequently used approach of employing non-parametric
multivariate analysis of variance (13). Regardless of these caveats,
McMurdie and Holmes recommended the use of variance stabilizing
transformations such as those applied for gene expression analysis such
as that found in the DeSeq R package.

Beyond the critiques from McMurdie and Holmes others have developed
alternative approaches to controlling for uneven sampling effort in
amplicon sequencing studies. For alpha diversity metrics, Willis used
toy datasets to demonstrate that one could estimate the richness for
each sample in a dataset and use those values for statistical
comparisons (14). Non-parametric estimators of richness and diversity
(15, 16) and parametric estimators of richness {[}Willis2015{]} that
have been used in microbial ecology studies. For beta diversity metrics
at least four approaches have been pursued. First, one could use
relative abundance values where the observed number of sequences in an
OTU is divided by the total number of sequences in the sample (17).
Second, normalization strategies have been developed where the relative
abundance is mutiplied by the size of the minimum desired sampling
effort and fractional values are reapportioned to the OTUs to obtain
integer values (18, 19). Third, a variety of center log-ratio methods
have been developed where the compositional nature of the OTU counts is
removed and used to calculated Euclidean distances (aka Aitchison
distances) {[}(20); (19); (17); (21); (22). This strategy is purported
to control for uneven sampling effort (21, 23); however, some have
noticed that this feature breaks down under certain conditions (24).
Finally, as mentioned above, variance stabilization transformations have
been recommended to generate values that can be used to calcualte
Euclidean distances (10).

The ongoing controversy over the use of rarefaction and the development
of numerous strategies to control for uneven sampling effort caused me
to question how these methods compared to each other. My analysis
included 16S rRNA gene sequence data from from 12 studies that
characterized the variation in bacterial communities from diverse
environments (Table 1 and Figure S1). The sequences were assigned to
OTUs using a standard pipeline and their frequencies and the number of
sequences found in each sample were used to generate simulated
communities and treatment effects. For each dataset and simulation, 100
replicate datasets were generated and used as input to each of the
strategies for controlling for uneven sampling effort. My overall
conclusion was that rarefaction outperformed the alternative strategies.

@@

Over the past decade, the question of whether to rarefy microbial
community sequence data has become controversial. The analyses I
presented here strongly indicate that rarefaction is necessary to
control for uneven sampling effort when comparing communities based on
alpha and beta diversity indices. Compared to all other methods,
rarefaction removed the correlation between sequencing depth and alpha
or beta diversity metrics when the sequencing depth varied by as much as
97-fold across samples. I showed that this correlation could lead
falsely detecting differences between treatment groups if sampling depth
and sequencing effort are confounded (e.g., 25). The correlation with
sampling effort leads to an artificial increase in the variation between
samples and a reduced power to detect true differences in alpha and beta
diversity. For these reasons, rarefaction is a valuable tool to control
for uneven sampling effort until improved statistical procedures are
developed or it becomes possible to more evenly distribute sequencing
effort across samples.

The primary alternative to rarefaction that many advise is to estimate
alpha diversity has been to use non-parametric or parametric methods
using raw data and to then compare the estimates (14). My data
demonstrates that such approaches are limited for several reasons.
First, it has been widely observed that non-parametric richness esimates
such as ACE and Chao1 are sensitive to sampling effort. Therefore, these
strategies do not, in practice, remove the effects of sampling effort.
Second, parametric approaches, such as those implemented in the
breakaway R package, generate confidence intervals that are likely to
include the true richness and theoretically shrink with increased
sampling effort. For most samples, the confidence intervals around the
estimates are too wide to be meaningful, again leading to an inability
to remvoe the effects of sampling effort. Third, it has become an
increasingly common practice for researchers to remove sequences that
are rare in a sample (e.g., those that appear once). Although that
approach was not taken in this study, removing rare sequences would skew
the distribtion of sequences and OTUs leading to a distortion of the
measurement of alpha diversity (12). The effects of removing rare
sequences would vary across samples depending on the number of sequences
in each sample. One interesting result of this analysis was the
demonstration that as metrics that depend less on rare taxa are used,
the effect of uneven sampling effort was reduced. For example, richness
counts a sequence appearing once as much as sequence appearing 1,000
times, while Shannon diversityindex places more emphasis on the more
abundant sequence, and the inverse Simpson index even more. Although
normalizing communities to a common number of sequences is also
suggested (e.g., SRS normalization) to control for uneven sampling
effort, the current analysis indicates that its performance does not
meet that of rarefaction. For analysis of alpha diversity metrics,
rarefaction is the most effective and consistent approach to control fo
uneven sampling effort.

Use of relative abundances, normalized counts, variance stabilizing
transformations, and centered log ratios have each been recommended as
more robust alternatives to rarefaction. Again, the only approach that
consistently removed the effects of uneven sampling effort was
rarefaction. Most of the recommendations borrow techniques from methods
for identifying differential gene expression. Unfortunately, there is an
important but underappreciated difference between gene expression and
community data. This is the interpretation of unobserved data. For gene
expression analysis in a single organism the lack of any sequencing data
for a gene would indicate that its expression was below the limit of
detection. Sequencing the same organism under multiple conditions would
not lead to a seemingly unbouned number of genes in the organims.
Rather, the number of genes has a definite limit that is knowable from
the genome sequence. With microbiome data, an unobserved sequence could
mean that the organism was present, but below the limit of detection or
that the organism was missing. Because we have yet to exhaustively
sample any community in the same way we have sequenced a single genome
it is unreliable to impute the presence of all organisms. Yet, this is
exactly what the variance stabilization transformation and most CLR
techniques do. This analysis has demonstrated a clear correlation
between distances calculated by these methods and sequencing effort.
This result is at odds with the claims by others that the distances are
scale invariant (21, 23). Again, rarefaction is the most effective and
consistent approach to control for uneven sampling effort when
calculating beta diversity metrics.

Two common critique of rarefaction is that the approach ommits valid
data and that the selection of the value to rarefy the sequence data to
is arbitrary (10). I disagree with both critiques. All of the data are
used to calculate the mean value of the rarefied metrics; none of it is
excluded. When the dataset is subsampled, every sequence has a random
chance of being included in the calculated metric. When that subsampling
is repeated a large number of times (e.g., 100 or 1000) the risks of
ignoring or oversampling rare taxa are mititgated. Furthermore, it is
curious that the study making the original critique removed sequences
that did not appear in at least three samples or had an abundance of one
in any sample. A parallel analysis to this study has demonstrated that
many of these sequences are likely valid and that removal of rare
sequences can bias alpha and beta diversity metrics and reduce
statistical power (12). As for the second critism, I would resist the
claim that the selection of the rarefaction threshold is arbitrary. In
practice, there is a tradeoff between sampling breadth and depth.
Greater breadth will increase the statistical power to compare treatment
groups and greater depth will increase the resolution to describe the
communities. My process for picking a break involves looking for a
natural break in the distribution of the number of sequences. For
example, the Lake dataset used in this study had a clear break around
10,000 sequences. I would also consider what samples are below any break
that I select. If there were critical samples below the break I would
either reduce the break point or I would generate more sequences from
those samples. Furthermore, regardless of the break point one selects,
they could repeat an analysis at multiple thresholds to assess the
tradeoff between statistical power and resolution. As shown in my
analysis of Good's coverage values, most studies obtain an ample level
of coverage and would need to increase their sampling depth by 10 fold
to increase the coverage by several percentage points. I contend that
for most studies increased sequencing effort should be applied to
improve sampling breadth and statistical power over depth and
resolution.

The up to 100-fold difference in sample sizes is an unfortunate
byproduct of how seqencing libraries are constructed. Researchers
perform separate PCRs for each sample with unique index (aka, barcode)
sequencs that all them to later identify which sample each sequence
belongs. When the PCRs are pooled efforts are taken to pool the
fragments in equimolar ratios. Researchers use one of two approaches.
First, they often will quantify the concentration of DNA from each PCR
and then pool DNA in the desired amounts. Alternatively, they may use
normalization plates where each well can hybridize a uniform amount of
DNA that is then eluted and pooled. Clearly, both approaches have
limitations that reduce the ability to truly achieve equimolar mixture.
In addition, for some samples it is common to co-amplify non-specific
DNAs which may add to the challenges of achieving equimolar mixtures of
the desired amplicons (25). Regardless, it is clear that better
strategies are needed to reduce the variation in the number of sequences
generated for each sample.

@@

\textbf{Choice of datasets.} The specific studies used in this study
were selected because their data was publicly accessible through the
Sequence Read Archive, the original investigators sequenced the V4
region of the 16S rRNA gene using paired 250 nt reads, and my previous
familiarity with the data. The use of paired 250 nt reads to sequence
the V4 region resulted in a near complete two-fold overap of the V4
region resulting in high quality contigs with a low sequencing error
rate (1). These data were processed through a standardized sequence
curation pipeline to generate operational taxonomic units (OTUs) using
the mothur software package (1, 8). OTUs were identified using the
OptiClust algorithm to cluster sequences together that were not more
than 3\% different from each other (26).

\textbf{Null community model.} Null community models were generated such
that within a dataset the number of sequences per sample and the number
of sequences per OTU across all samples within the dataset were the same
as was observed in original. This model effectively generated
statistical samples of a single community so that there should not have
been a statistical difference between the samples. This model
implemented by randomly assigning each sequence in the dataset to an OTU
and sample while keeping constant the number of sequences per sample and
the total number of sequences in each OTU. This is a similar approach to
that of the IS algorithm described by Ulrich and Gotelli (27). Because
the construction of the null models was a stochastic process, 100
replicates were geneated for each dataset.

@@

\textbf{Test of statistical significance.} Statistical comparisons of
alpha diversity metrics across the simulated treatment groups were
performed using the non-parametric two-sample Wilcoxon test as
implemented in R. This test was selected because the alpha-diversity
metrics tended to not be normally distributed and each dataset required
a different transformation to normalize the data. Comparisons of beta
diversity metrics were performed using the adonis2 function from the
vegan (v.2.6.2) R package (9). The adonis2 function implements a
non-parametric multivariate analysis of variance using distances
matrices as described by McArdle and Anderson (13). Throughout this
study I used 0.05 as the threshold for asssessing statistical
significance.

@@

\textbf{Alpha diversity calculations.} Various strategies for handling
uneven sampling effort were evaluated to identify the best approach for
calculating community richness and Shannon and inverse Simpson diversity
indeices. Raw OTU counts were used as input to calculate sample richness
and Shannon and inverse Simpson diversity using mothur (8, 28). Shannon
diversity was calculated as

@@

The inverse Simpson diversity was calculated as \(1/D_{simpson}\). In
both formulae, \(n_i\) is the number of sequences in OTU \(i\) and \(N\)
is the number of sequences in the sample. Rarefaction of richness,
Shannon diversity and Inverse Simpson diversity values were carried out
in mothur. Briefly, mothur calculates each value on a random draw of the
same number of sequences from each sample and obtains a mean value based
on 1,000 random draws. Scaled ranked subsampling (SRS) was used to
normalize OTU counts to the size of the smallest sample in each dataset
using the SRS R package (v.0.2.3)(18). Normalized OTU counts were used
to calculate sample richness and Shannon and inverse Simpson diversity
values using mothur. Data normalized by cumulative sum scaling (CSS)
were not reported for alpha-diversity values since the relative
abundances of the features do not change with the normalization
procedure (19). The non-parametric bias-corrected Chao1 and ACE richness
estimators (16) and a non-parametric estimator of the Shannon diversity
(15) were calculated using raw OTU counts with mothur. Parametric
estimates of sample richness were calculated using the breakaway (BA) R
package (v.4.7.9)(29). The current analysis reports both the results
from running default model selection procedure and the Poisson model.
The default model selection returned either the Kemp, Negative Binomial,
or Poisson models. Relative abundance data were not used to calculate
alpha diversity metrics since the richness and evenness does not change
from the raw data when dividing each sample by the total number of
sequences in the sample.

\textbf{Beta diversity calculations.} Similar to the alpha diversity
calculations, multiple approaches were used to control for uneven
sampling effort and calculate beta diversity. Raw and OTU counts were
used for input to calculate the Jaccard, Bray-Curtis, and Euclidean
dissimilarity indices using the vegdist function from the vegan R
package (v.2.6.2)(9). The Jaccard index was calculated as

@@

Three approaches were taken to normalize the number of sequences across
samples within a dataset. Scaled ranked subsampling (SRS) and cumulative
sum scaling (CSS) were used to normalize raw OTU counts using the SRS
(v.0.2.3) and metagenomeSeq (v.1.36.0) R packages, respectively (18,
19). The normalized counts were then used to calculate Jaccard and
Bray-Curtis dissimilarity indices. Finally, the variance-stabilization
transformation (VST) as implemented in the DESeq2 (v.1.34.0) R package
was used to normalize the data as described by McMurdie and Holmes (10,
30). Because the VST approach generates negative values, which are
incompatible with calculating Jaccard and Bray-Curtis dissimilarity
values, Euclidean distances were calcualted instead.

@@

where \$x\_\{ij\} was the number of sequences observed for OTU \(i\) in
sample \(j\) and \(g()\) was the geometric mean \(x_j\) was the counts
of the \(S_T\) OTUs in sample \(j\). Because the geometric mean is zero
if any OTU is absent from a sample, the CLR is undefined when there are
unobserved OTUs in a sample. To overcome this problem, I attempted a
four approaches. The first, Zero CLR, imputed the value of the zeroes
based on the observed data using the zCompositions (v.1.4.0.1) R package
(31). The second, One CLR, added a pseudocount of 1 to the abundance of
all OTUs (17, 19). The third, Nudge CLR, added a pseudocount of 1
divided by the total number of sequences in a sample to each OTU in the
sample (17, 22). The final approach, Robust CLR, removed unobserved OTUs
prior to calcualting the CLR (21).

@@

\hypertarget{refs}{}
\begin{CSLReferences}{0}{1}
\leavevmode\vadjust pre{\hypertarget{ref-Kozich2013}{}}%
\CSLLeftMargin{1. }%
\CSLRightInline{\textbf{Kozich JJ}, \textbf{Westcott SL}, \textbf{Baxter
NT}, \textbf{Highlander SK}, \textbf{Schloss PD}. 2013. {Development of
a dual-index sequencing strategy and curation pipeline for analyzing
amplicon sequence data on the MiSeq Illumina sequencing platform}.
Applied and environmental microbiology \textbf{79}:5112--5120.}

\leavevmode\vadjust pre{\hypertarget{ref-Caporaso2010}{}}%
\CSLLeftMargin{2. }%
\CSLRightInline{\textbf{Caporaso JG}, \textbf{Lauber CL},
\textbf{Walters WA}, \textbf{Berg-Lyons D}, \textbf{Lozupone CA},
\textbf{Turnbaugh PJ}, \textbf{Fierer N}, \textbf{Knight R}. 2010.
Global patterns of 16S {rRNA} diversity at a depth of millions of
sequences per sample. Proceedings of the National Academy of Sciences
\textbf{108}:4516--4522.
doi:\href{https://doi.org/10.1073/pnas.1000080107}{10.1073/pnas.1000080107}.}

\leavevmode\vadjust pre{\hypertarget{ref-Sanders1968}{}}%
\CSLLeftMargin{3. }%
\CSLRightInline{\textbf{Sanders HL}. 1968. Marine benthic diversity: A
comparative study. The American Naturalist \textbf{102}:243--282.
doi:\href{https://doi.org/10.1086/282541}{10.1086/282541}.}

\leavevmode\vadjust pre{\hypertarget{ref-Hurlbert1971}{}}%
\CSLLeftMargin{4. }%
\CSLRightInline{\textbf{Hurlbert SH}. 1971. The nonconcept of species
diversity: A critique and alternative parameters. Ecology
\textbf{52}:577--586.
doi:\href{https://doi.org/10.2307/1934145}{10.2307/1934145}.}

\leavevmode\vadjust pre{\hypertarget{ref-Dunbar1999}{}}%
\CSLLeftMargin{5. }%
\CSLRightInline{\textbf{Dunbar J}, \textbf{Takala S}, \textbf{Barns SM},
\textbf{Davis JA}, \textbf{Kuske CR}. 1999. Levels of bacterial
community diversity in four arid soils compared by cultivation and 16S
{rRNA} gene cloning. Applied and Environmental Microbiology
\textbf{65}:1662--1669.
doi:\href{https://doi.org/10.1128/aem.65.4.1662-1669.1999}{10.1128/aem.65.4.1662-1669.1999}.}
